---
title: "Flipper length and body mass in three penguin species"
subtitle: "A minimal reproducible report"
author: "Your name"
date: today
format:
  html:
    toc: true
    embed-resources: true
    fig-width: 6
    fig-height: 3.6
execute:
  echo: true
  warning: false
  message: false
---

<!-- Basics in Statistics, Unit 4: Press "Render" in RStudio. -->

```{r}
#| label: setup
# install.packages(c("palmerpenguins", "dplyr", "ggplot2", "broom", "knitr",
#                    "rmarkdown"))
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
```

## Data

The `penguins` data set records body measurements of adult foraging penguins
near Palmer Station, Antarctica. We use flipper length and body mass and drop
the rows in which one of them is missing, stating how many that is.

```{r}
#| label: data
penguins_complete <- penguins |>
  filter(!is.na(flipper_length_mm), !is.na(body_mass_g))

n_all <- nrow(penguins)
n_used <- nrow(penguins_complete)
```

Of the `r n_all` birds in the data set, `r n_used` have complete measurements
and are used below; `r n_all - n_used` are dropped.

```{r}
#| label: tbl-summary
#| tbl-cap: "Body mass and flipper length by species, for the birds used here."
penguins_complete |>
  group_by(Species = species) |>
  summarise(
    n = n(),
    `Mean body mass (g)` = round(mean(body_mass_g)),
    `SD body mass (g)` = round(sd(body_mass_g)),
    `Mean flipper (mm)` = round(mean(flipper_length_mm), 1),
    .groups = "drop"
  ) |>
  kable()
```

## A first look

```{r}
#| label: fig-scatter
#| fig-cap: "Body mass against flipper length. Each point is one bird."
ggplot(penguins_complete,
       aes(flipper_length_mm, body_mass_g, shape = species, colour = species)) +
  geom_point(alpha = 0.75, size = 1.8) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", legend.title = element_blank())
```

@fig-scatter shows a clear positive association, and it also shows that the
three species sit in different places along both axes. A model that ignores
species would therefore mix two things: How mass changes with flipper length
within a species, and how the species differ.

## Model

We regress body mass on flipper length and species. The species coefficients are
differences from the reference species, Adelie.

```{r}
#| label: model
fit <- lm(body_mass_g ~ flipper_length_mm + species, data = penguins_complete)
```

```{r}
#| label: tbl-coefs
#| tbl-cap: "Estimated coefficients with 95% confidence intervals."
tidy(fit, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)),
         p.value = format.pval(p.value, eps = 0.001)) |>
  select(Term = term, Estimate = estimate, `Std. error` = std.error,
         `CI lower` = conf.low, `CI upper` = conf.high, `p value` = p.value) |>
  kable()
```

```{r}
#| label: slope
slope <- tidy(fit, conf.int = TRUE) |> filter(term == "flipper_length_mm")
```

Within a species, a flipper that is one millimetre longer goes with a body mass
that is on average `r round(slope$estimate, 1)` g larger
(95% CI `r round(slope$conf.low, 1)` to `r round(slope$conf.high, 1)` g).
The model reproduces `r round(100 * summary(fit)$r.squared)`% of the variation in
body mass.

## What this report does not claim

The birds were measured, not assigned to a flipper length. The estimate
describes an association within these three species at this location; it is not
an estimate of what would happen to a bird's mass if its flipper grew.

## Reproducibility

```{r}
#| label: session
sessionInfo()
```
